---
title: "SimLGCP"
format: pdf
editor: visual
---

## Package Installation

```{r}
library(mgcv)

library(spatstat)
```

## Simulating LGCP Data Using SpatStat

### Define the spatial window

Using unit square for this example

```{r}
win <- owin(xrange = c(0,1), yrange = c(0,1))
```

### Define Covariate Functions

```{r}
cov_x <- as.im(function(x,y) x, W = win)
cov_y <- as.im(function(x,y) y, W = win)
cov_z <- as.im(function(x,y) x+y+2, W = win)
cov_a <- as.im(function(x,y) 2*x, W = win)
```

### Define Mean Function Based on Covariates

```{r}
mean <- function(x,y)
  {2*x + x + y + 2}

mean.im <- as.im(mean, W = win)
```

### Simulate LGCP using rLGCP()

#### Exponential

```{r}
set.seed(111)
ppp_exp <- rLGCP("exponential", mu = mean.im, var = 1, scale = 0.05, win = win)

#Change var
ppp_exp_var1 <- rLGCP("exponential", mu = mean.im, var = 2, scale = 0.05, win = win)

ppp_exp_var2 <- rLGCP("exponential", mu = mean.im, var = 0.5, scale = 0.05, win = win)

# Change scale
ppp_exp_scale1 <- rLGCP("exponential", mu = mean.im, var = 1, scale = 0.09, win = win)
# if scale goes above 1, becomes similar to small scale

ppp_exp_scale2 <- rLGCP("exponential", mu = mean.im, var = 1, scale = 0.001, win = win)

par(mfrow = c(2, 3), mar = c(2, 2, 2, 1)) 

plot(ppp_exp, main = "Default: var=1, scale=0.05", pch = 16, cex = 0.5)
plot(ppp_exp_var1, main = "Var = 2", pch = 16, cex = 0.5)
plot(ppp_exp_var2, main = "Var = 0.5", pch = 16, cex = 0.5)
plot(ppp_exp_scale1, main = "Scale = 0.09", pch = 16, cex = 0.5)
plot(ppp_exp_scale2, main = "Scale = 0.001", pch = 16, cex = 0.5)


```

#### Gauss

```{r}
set.seed(112)
ppp_gauss <- rLGCP("gauss", mu = mean.im, var = 1, scale = 0.05, win = win)

#Change var
ppp_gauss_var1 <- rLGCP("gauss", mu = mean.im, var = 2, scale = 0.05, win = win)

ppp_gauss_var2 <- rLGCP("gauss", mu = mean.im, var = 0.5, scale = 0.05, win = win)

# Change scale
ppp_gauss_scale1 <- rLGCP("gauss", mu = mean.im, var = 1, scale = 0.09, win = win)
# if scale goes above 1, becomes similar to small scale

ppp_gauss_scale2 <- rLGCP("gauss", mu = mean.im, var = 1, scale = 0.001, win = win)

par(mfrow = c(2, 3), mar = c(2, 2, 2, 1))  
plot(ppp_gauss, main = "Default: var=1, scale=0.05", pch = 16, cex = 0.5)
plot(ppp_gauss_var1, main = "Var = 2", pch = 16, cex = 0.5)
plot(ppp_gauss_var2, main = "Var = 0.5", pch = 16, cex = 0.5)
plot(ppp_gauss_scale1, main = "Scale = 0.09", pch = 16, cex = 0.5)
plot(ppp_gauss_scale2, main = "Scale = 0.001", pch = 16, cex = 0.5)
```

#### Stable

```{r}
set.seed(113)

ppp_stable <- rLGCP("stable", mu = mean.im, var = 1, scale = 0.05, alpha = 0.5, win = win)

#Change var
ppp_stable_var1 <- rLGCP("stable", mu = mean.im, var = 2, scale = 0.05,alpha = 0.5, win = win)

ppp_stable_var2 <- rLGCP("stable", mu = mean.im, var = 0.5, scale = 0.05, alpha = 0.5,win = win)

# Change scale
ppp_stable_scale1 <- rLGCP("stable", mu = mean.im, var = 1, scale = 5, alpha = 0.5,win = win)

ppp_stable_scale2 <- rLGCP("stable", mu = mean.im, var = 1, scale = 0.001, alpha = 0.5,win = win)

# Change alpha
ppp_stable_alpha1 <- rLGCP("stable", mu = mean.im, var = 1, scale = 0.001, alpha = 5,win = win)

ppp_stable_alpha2 <- rLGCP("stable", mu = mean.im, var = 1, scale = 0.001, alpha = 0.01,win = win)

par(mfrow = c(3, 3), mar = c(2, 2, 2, 1)) 

plot(ppp_stable, main = "Default: var=1, scale=0.05, α=0.5", pch = 16, cex = 0.5)
plot(ppp_stable_var1, main = "Var = 2", pch = 16, cex = 0.5)
plot(ppp_stable_var2, main = "Var = 0.5", pch = 16, cex = 0.5)
plot(ppp_stable_scale1, main = "Scale = 5", pch = 16, cex = 0.5)
plot(ppp_stable_scale2, main = "Scale = 0.001", pch = 16, cex = 0.5)
plot(ppp_stable_alpha1, main = "Alpha = 5", pch = 16, cex = 0.5)
plot(ppp_stable_alpha2, main = "Alpha = 0.01", pch = 16, cex = 0.5)


```

#### Generalised Cauchy

```{r}
set.seed(114)

ppp_gencauchy <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.05, alpha = 1, beta = 1, win = win)

#Change var
ppp_gencauchy_var1 <- rLGCP("gencauchy", mu = mean.im, var = 2, scale = 0.05, alpha = 1, beta = 1, win = win)

ppp_gencauchy_var2 <- rLGCP("gencauchy", mu = mean.im, var = 0.5, scale = 0.05, alpha = 1, beta = 1, win = win)

#Change scale
ppp_gencauchy_scale1 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 5, alpha = 1, beta = 1, win = win)

ppp_gencauchy_scale2 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.001, alpha = 1, beta = 1, win = win)

#Change alpha
ppp_gencauchy_alpha1 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.05, alpha = 5, beta = 1, win = win)

ppp_gencauchy_alpha2 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.05, alpha = 0.01, beta = 1, win = win)

#Change beta
ppp_gencauchy_beta1 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.05, alpha = 1, beta = 5, win = win)

ppp_gencauchy_beta2 <- rLGCP("gencauchy", mu = mean.im, var = 1, scale = 0.05, alpha = 1, beta = 0.01, win = win)

par(mfrow = c(3, 3), mar = c(2, 2, 2, 1))

plot(ppp_gencauchy, main = "Default: var=1, scale=0.05", pch = 16, cex = 0.5)
plot(ppp_gencauchy_var1, main = "Var = 2", pch = 16, cex = 0.5)
plot(ppp_gencauchy_var2, main = "Var = 0.5", pch = 16, cex = 0.5)
plot(ppp_gencauchy_scale1, main = "Scale = 5", pch = 16, cex = 0.5)
plot(ppp_gencauchy_scale2, main = "Scale = 0.001", pch = 16, cex = 0.5)
plot(ppp_gencauchy_alpha1, main = "Alpha = 5", pch = 16, cex = 0.5)
plot(ppp_gencauchy_alpha2, main = "Alpha = 0.01", pch = 16, cex = 0.5)
plot(ppp_gencauchy_beta1, main = "Beta = 5", pch = 16, cex = 0.5)
plot(ppp_gencauchy_beta2, main = "Beta = 0.01", pch = 16, cex = 0.5)

```

#### Matern

```{r}
set.seed(115)

ppp_matern <- rLGCP("matern", mu = mean.im, var = 1, scale = 0.5, nu = 0.05, win = win)

#Change var
ppp_matern_var1 <- rLGCP("matern", mu = mean.im, var = 2, scale = 0.5, nu = 0.05, win = win)

ppp_matern_var2 <- rLGCP("matern", mu = mean.im, var = 0.5, scale = 0.5, nu = 0.05, win = win)

#Change scale
ppp_matern_scale1 <- rLGCP("matern", mu = mean.im, var = 1, scale = 5, nu = 0.05, win = win)

ppp_matern_scale2 <- rLGCP("matern", mu = mean.im, var = 1, scale = 0.001, nu = 0.05, win = win)

#Change nu
ppp_matern_nu1 <- rLGCP("matern", mu = mean.im, var = 1, scale = 0.5, nu = 5, win = win)

ppp_matern_nu2 <- rLGCP("matern", mu = mean.im, var = 1, scale = 0.5, nu = 0.001, win = win)

par(mfrow = c(3, 3), mar = c(2, 2, 2, 1)) 


plot(ppp_matern, main = "Default: var=1, scale=0.5, nu=0.05", pch = 16, cex = 0.5)
plot(ppp_matern_var1, main = "Var = 2", pch = 16, cex = 0.5)
plot(ppp_matern_var2, main = "Var = 0.5", pch = 16, cex = 0.5)
plot(ppp_matern_scale1, main = "Scale = 5", pch = 16, cex = 0.5)
plot(ppp_matern_scale2, main = "Scale = 0.001", pch = 16, cex = 0.5)
plot(ppp_matern_nu1, main = "Nu = 5", pch = 16, cex = 0.5)
plot(ppp_matern_nu2, main = "Nu = 0.001", pch = 16, cex = 0.5)
```

### Simulate Covariate Values Using Defined Functions for Observed Points

Using Gaussian model for rest of fitting (but can use any)

```{r}
lgcp_df_gauss <- as.data.frame(ppp_gauss)

lgcp_df_gauss$cov_x <- cov_x[lgcp_df_gauss]
lgcp_df_gauss$cov_y <- cov_y[lgcp_df_gauss]
lgcp_df_gauss$cov_z <- cov_z[lgcp_df_gauss]
lgcp_df_gauss$cov_a <- cov_a[lgcp_df_gauss]

# Adding pt and wt for the "observed points"
lgcp_df_gauss$pt <- 1
lgcp_df_gauss$wt <- 1e-6
```

### Create Grid in Window for Quadrature Points

```{r}
# mask <- as.mask(win, eps = 0.01)

# dummy_points <- gridcentres(mask)

quad_points <- quadscheme(ppp_gauss, method = "grid")

plot(quad_points$data)
plot(quad_points$dummy)
```

### Create Quadrature Data Frame

```{r}

quad_df_gauss <- data.frame(
  x = quad_points$dummy$x, 
  y = quad_points$dummy$y,
  pt = 0,
  wt = 6.94e-05 # came from areas vector in quad_points list
)

```

### Generate Quadrature Covariates

```{r}
quad_df_gauss$cov_x <- cov_x[quad_df_gauss]
quad_df_gauss$cov_y <- cov_y[quad_df_gauss]
quad_df_gauss$cov_z <- cov_z[quad_df_gauss]
quad_df_gauss$cov_a <- cov_a[quad_df_gauss]
```

### Use rbind() to Merge Observed and Quadrature Points Together

```{r}
data_gauss <- rbind(lgcp_df_gauss, quad_df_gauss)

data_gauss
```
## K Ratio Comparisons
From GAM paper - "Wood(2017, sec. 5.9) suggests that k should be large relative to the empirical degrees of freedom, computed as the trace of the influence matrix of the fitted model, and so we chose k in our applications such that it was at least four times larger than the empirical degrees of freedom (see supplementary material Section S1.3 for further on this)."

### Create function to look at different basis function amounts
```{r}
basis_func <- function(data, k, m = "3", 
                       response = "pt", 
                       covariates = c("cov_x", "cov_y", "cov_z"),
                       family = "poisson()",
                       weight = "wt",
                       method = "REML",
                       bs = "gp"){
  
  linear <- paste0(covariates, collapse = "+")
  
  smoothing <- paste0("s(x, y, bs = '", bs, "', k = ", k, ", m = ", m, ")")
  
  # Combining both sides
  
  rhs <- paste0(linear, "+", smoothing)
  formula_str <- paste0(response, " / ", weight, " ~ ", rhs)
  
  
  # Convert to formula
  model_formula <- as.formula(formula_str)
  
  # Fit the GAM model
  gam_model <- gam(model_formula,
      data = data,
      family = poisson(),
      weights = data[[weight]],
      method = "REML")
  
  # K ratio
  
  k_ratio <- sum(gam_model$edf) / k
  
  result <- list("Model" = gam_model, "K Ratio" = k_ratio)
  
  return(result)
  
}

```

### Comparing k-ratios 
```{r}

basis_func(data = data_gauss, k = 30)

basis_func(data = data_gauss, k = 40)

basis_func(data = data_gauss, k = 50)

basis_func(data = data_gauss, k = 60)

basis_func(data = data_gauss, k = 70)


```
k = 70 seems like the best as it has a k ratio of 0.24 which reaches the threshold of 1/4 the edf of the model

## Optimizing rho (range parameter, m)

Found out they provide a wrapper function for optimizing called "gam_lgcp.R"

```{r}
source("gam_lgcp.R")

m_k70 <- basis_func(data = data_gauss, k = 70)

gam_lgcp(m_k70, data = data_gauss, weights = wt)
```




